Principles for Open Data Curation: A Case Study with the New York City 311 Service Request Data

### 2024-10-21 manuscript/manuscript.tex
Further editing

Collectively these six City agnecies handle over 90\% of the total SR 
volume, indicative of the critical role these agencies play in managing
public concerns, ranging from noise complaints and housing issues to
sanitation and transportation problems. The NYPD alone accounts for a
substantial portion of the total SRs, underscoring its leading role in
addressing complaints related to public safety. The remaining
10\% of SRs is distributed across 10 other
agencies. This concentration of SRs to the `big six'' agencies 
underscores the necessity for optimal data handling and curation processes 
within these high-volume agencies in order to ensure 
efficiency and responsiveness.

@@

Data cleansing refers to the process of identifying and rectifying
errors, inconsistencies, and inaccuracies within datasets to ensure
they are of high quality and reliable for analysis
\citep{maletic2005data, hosseinzadeh2023data}. The process
typically involves removing duplicate records, handling missing or
incomplete data, correcting mislabeled or inaccurate entries, and
standardizing data formats \citep[e.g.,][]{cody2017cody,
  van2018statistical}. In the context of open data, cleansing is
especially important as open datasets often come from diverse,
uncoordinated sources, leading to variations in data quality,
completeness, and consistency. Without thorough data cleansing, the utility
of open data can be limited and perhaps untrustworhty, affecting 
its reliability for research, policy-making, and innovation. The 
main purpose of cleansing open data is to ensure that it is 
accurate, consistent, and usable across multiple platforms and by 
various stakeholders with a myriad of purpose. Data cleansing improves
the trustworthiness of the data and enables more accurate analysis, 
better decision-making,and the integration of data into machine
learning models or other systems.

@@

	\item There are four date fields (created, closed, action\_updated, due).

@@

	\item In addition to the incident\_address, there are five additional location fields: 
	latitude/longitude, street\_name, landmark, USGS State Plane 
	Coordinate System, and Block \#.

	\item A free-form text field, resolution\_description, which 
	supports 934 characters of input, including commas and special characters
\end{itemize}

@@

The Mostly Empty category includes such fields as
\texttt{taxi\_company due\_date},
\texttt{pickup\_location}, and \texttt{landmark}.
The Partially Empty includes such fields as
\texttt{location\_type}, \texttt{borough}, 
and \texttt{cross\_street}. And the Few/None Empty includes
\texttt{created\_date}, \texttt{complaint\_type},
\texttt{agency}, and \texttt{status}. In some analysis efforts, it may be
prudent to inquire as to why some fields are mostly or almost always blank.
Figure~\ref{fig:blank_fields} presents a graphic 
depiction of total empty (blank \& N/As) for each fields illustrating
the grouping into the Most, Partial, and Few categories.

@@

\paragraph{Zip Codes}
\label{sec:zipcodesissues}
All zip codes ( \texttt{incident\_zip}) 
should validate against the USPS database, which contains 
44,173 valid zip codes. The \texttt{incident\_zip} field has 
only 0.07\% invalid entries, this still results in 4,374 errors; 
not insignificant. If you group the invalid entries by 
percentage by Agency, it mirrors the \texttt{overall} 
breakdown of SRs by Agency, potentially 
indicating a systemic problem.

\paragraph{Created and Closed Date}
\label{sec:negativeduration}
For the \texttt{created\_date} and \texttt{closed\_date} fields, one 
might expect that these fields are automatically populated by the SR 
application software when setting an SR status to ``new'' or ``closed.'' 
Unfortunately, this does not seem to be the case, as several anomalies 
exist in the created and closed date fields. Here are some of the errors:

\begin{itemize}
    \item SRs with a \texttt{closed\_date} that occurs before the 
    \texttt{created\_date} indicating a negative duration.
    \item \texttt{created\_date(s)} and \texttt{closed\_date(s)} in 
    the far distant past, e.g. 1900.
    \item \texttt{created\_date(s)} and \texttt{closed\_date(s)} that 
    match to the second indicating a zero duration.
    \item A large number of SRs closed and/or created exactly at midnight 
    or noon, to the second.
\end{itemize}

\subparagraph{Closed before Created: Negative Duration}
Citizens, NYC Government Officials, and Agencies use the created and 
closed dates to measure the ``duration'' of SRs, which reflects Agency 
responsiveness. While duration is not directly present in the dataset, 
it can be easily computed as the difference between
\texttt{(closed\_date} and \texttt{created\_date}.  Duration is one of 
most used fields that is constantly analyzed, such as determining if 
one NYC borough receives faster response for ceratin 
complaints. There are 12,450 SRs where the \texttt{closed\_date} precedes the 
\texttt{created\_date}, generating nonsensical ``negative durations.'' 
Though this represents only 0.2\% of the dataset, these errors can 
significantly impact response time analyses. Eight SRs with extremely 
large negative durations ($-$730 days), all originating from the 
Department of Homeless Services (DHS); all contain an entry 
of ``1900-01-01'' as the \texttt{closed\_date}. This results 
in negative durations exceeding $-$44,601 days (or 122 years). Many
other SRs have negative durations in excess of 300+ days. Such 
anomalies, though rare, can significantly skew statistical results if 
not addressed during data cleansing. As a result, these SR rows are 
removed from the analysis. Here is a sample:

\begin{table}[tbp]
  \centering
  \caption{Negative duraton of the largest magnitude (days) excluding
    extreme negative values}
  \begin{tabular}{l l l r l}
    \toprule
    {created\_date} & {closed\_date} & {duration} 
    & \textbf{agency} \\
    \midrule
    2023-01-27 14:40:00 & 2022-01-14 14:40:00 & $-$378.0 & DOT \\
    2023-01-18 10:06:00 & 2022-01-12 10:06:00 & $-$371.0 & DOT \\
    2023-01-27 14:36:00 & 2022-01-22 14:35:00 & $-$370.0 & DOT \\
    2023-01-11 11:10:00 & 2022-01-09 11:10:00 & $-$367.0 & DOT \\
    2023-12-18 03:13:00 & 2023-01-16 13:10:00 & $-$335.6 & DOT \\
    \bottomrule
  \end{tabular}
  \label{tab:largest-errors}
\end{table}

@@

\subparagraph{Resolution Action Update Date}

A similar issue occurs when resolution action update date is after the
closing date.  When an SR is updated, the 311 software 
automatically populates the \texttt{resolution\_action\_update\_date}. 
While it is certainly possible to update an SR after it is closed, 
some of these updates seem to be well beyond an expected 
range, some over 2 years later. As shown in 
Figure~\ref{fig:resolution-violin}, there are 10,265 SRs that 
were updated more than 30 days after the 
\texttt{closed\_date}, but less than 730 days, cutoff used to 
exclude infeasible dates such as 1900-01-01. The chart 
highlights the distribution of these post-closure updates, with 
a notable concentration of updates within 
the 30-to-90-day range. This pattern raises questions about whether 
such delayed updates are standard operational practice or indicative 
of a potential issue requiring further investigation.

\begin{figure}[tbp]
  \centering
  \includegraphics[width=\textwidth]{post_closed_violin_chart.pdf}
  \caption{Post-closed resolution action update date(s) over 30 days.}
  \label{fig:resolution-violin}
\end{figure}

\subparagraph{Identical Created Date \& Closed Date: Zero Durations}
A more prevalent issue occurs when the \texttt{closed\_date} and 
\texttt{created\_date} are exactly the same, down to the second. This 
creates a \textbf{zero duration}, which is nonsensical. There 
are 163,720 such SRs, representing 2.6\% of all non-blank data. 
Ninety-nine percent of these zero-duration SRs occur in five agencies:
DHMH, DOT, DOB, DSNY, and DEP. This distribution does not 
mirror the overall SR distribution by Agency, suggesting 
an agency-specific issue.

@@

These unusual patterns of SR creation and closure exactly at midnight 
and noon suggest the presence of a bulk create/close software process 
that automatically assigns time stamps of midnight (00:00:00) or noon 
(12:00:00) to large batches of SRs. This behavior distorts the 
calculated duration of these SRs by providing inaccurate dates. The 
distribution by agency shows that over 90\% of the ``created-at-noon'' 
SRs come from just two agencies: DOB and DSNY. Further investigation 
reveals that a single agency, DSNY, is responsible for over 99\% of 
the ``closed-at-noon'' SRs.

@@

An question of precision vs. accuracy arises with the Latitude 
and Longitude fields. Both Latitude are expressed as 
a 14-decimal number, e.g. 40.86769186022511. Given 
that 1 degree of latitude at the equator is equal to 111.04 
kilometers, the ``1'' at the end of that number represents 
approximately 1.1 nanometers (1/1,000,000,000 of a meter). For 
reference a DNA molecule is approximately 2nm in width. The 
representation of the Latitude and Longitude fields appears to be a 
classic case of 14-digit precision, but limited accuracy.

@@

\textbf{Latitude/Longitude and Location:} The \texttt{location} field 
is a pure concatenation of the latitude and longitude fields with a 
comma and parentheses added. This makes the \texttt{location} field 
arguably harder to use than the individual fields. For example:  
\texttt{latitude}: 40.768456429488, \texttt{longitude}: $-$73.9575661888774, 
and \texttt{location}: (40.768456429488, $-$73.95756618887745).

\textbf{borough \& park\_borough:} These two fields are fully redundant, 
as they are 100\% matches.

\textbf{borough \& taxi\_company\_borough:} Despite their names, these 
fields are almost entirely different, with only a 0.05\% match. The 
\texttt{taxi\_company\_borough} field is used exclusively by the TLC, 
indicating a need for consultation with TLC to understand the differing 
uses.
\jy{The taxi\_company\_borough may mean the borough of the location of
  the taxi company, which can be irrelevant to the borough where the
  incident was reported.. Such information could be useful only when
  studying TLC related SRs. It'd be better to use a second table to
  store such TLC dependent information.}

>>>>

Collectively these six City agencies handle over 90\% of the total SR 
volume, indicative of the critical role these agencies play in managing
public concerns, ranging from noise complaints and housing issues to
sanitation and transportation problems. The NYPD alone accounts for a
substantial portion of the total SRs, underscoring its leading role in
addressing complaints related to public safety. The remaining
10\% of SRs is distributed across 10 other
agencies. This concentration of SRs to the `big six'' agencies 
underscores the necessity for optimal data handling and curation processes 
within these high-volume agencies in order to ensure 
efficiency and responsiveness.

@@

Data cleansing refers to the process of identifying and rectifying
errors, inconsistencies, and inaccuracies within datasets to ensure
they are of high quality and reliable for analysis
\citep{maletic2005data, hosseinzadeh2023data}. The process
typically involves removing duplicate records, handling missing or
incomplete data, correcting mislabeled or inaccurate entries, and
standardizing data formats \citep[e.g.,][]{cody2017cody,
  van2018statistical}. In the context of open data, cleansing is
especially important as open datasets often come from diverse,
uncoordinated sources, leading to variations in data quality,
completeness, and consistency. Without thorough data cleansing, the utility
of open data can be limited and perhaps untrustworthy, affecting 
its reliability for research, policy-making, and innovation. The 
main purpose of cleansing open data is to ensure that it is 
accurate, consistent, and usable across multiple platforms and by 
various stakeholders with a myriad of purpose. Data cleansing improves
the trustworthiness of the data and enables more accurate analysis, 
better decision-making,and the integration of data into machine
learning models or other systems.

@@

	\item There are four date fields 
	(\texttt{created\_date}, \texttt{closed\_date}, \texttt{resolution\_action\_updated\_date}, \& \texttt{due\_date}).

@@

	\item In addition to  \texttt{incident\_address}, there are five additional location fields: 
	\texttt{latitude \& longitude}, \texttt{street\_name}, \texttt{landmark}, USGS State Plane 
	Coordinate System, and \texttt{block} \#.

	\item A free-form text field, \texttt{resolution\_description}, which 
	supports 934 characters of input, including commas and special characters
\end{itemize}

@@

The Mostly Empty category includes such fields as
\texttt{taxi\_company due\_date},
\texttt{pickup\_location}, and \texttt{landmark}.
The Partially Empty includes such fields as
\texttt{location\_type}, \texttt{borough}, 
and \texttt{cross\_street\_1 \& \_2}. And the Few/None Empty includes
\texttt{created\_date}, \texttt{complaint\_type},
\texttt{agency}, and \texttt{status}. In some analysis efforts, it may be
prudent to inquire as to why some fields are mostly or almost always blank.
Figure~\ref{fig:blank_fields} presents a graphic 
depiction of total empty (blank \& N/As) for each fields illustrating
the grouping into the Most, Partial, and Few categories.

@@

\paragraph{Zip Codes}
\label{sec:zipcodesissues}
All zip codes ( \texttt{incident\_zip}) 
should validate against the USPS database, which contains 
44,173 valid zip codes. The \texttt{incident\_zip} field has 
only 0.07\% invalid entries, this still results in 4,374 errors; 
not insignificant. If you group the invalid entries by 
percentage by Agency, it mirrors the \textit{overall} 
breakdown of SRs by Agency, potentially 
indicating a systemic problem.

\paragraph{Created and Closed Date}
\label{sec:negativeduration}
For the \texttt{created\_date} and \texttt{closed\_date} fields, one 
might expect that these fields to be automatically populated by the  
application software when setting an SR status to ``new'' or ``closed.'' 
Unfortunately, this does not seem to be the case, as several anomalies 
exist in the \texttt{created\_date} and \texttt{closed\_date} fields that
an automated process would normally exclude, including:

\begin{itemize}
    \item SRs with a \texttt{closed\_date} that occurs before the 
    \texttt{created\_date} creating a ``negative duration''.
    \item \texttt{created\_date(s)} and \texttt{closed\_date(s)} in 
    the far distant past, e.g. ``1900-01-01''.
    \item \texttt{created\_date(s)} and \texttt{closed\_date(s)} that 
    match \textbf{to the second} indicating a nonsensical ``zero duration''.
    \item An unusually large number of SRs closed and/or created exactly at midnight 
    and noon, to the second.
\end{itemize}

\subparagraph{Closed before Created: Negative Duration}
Citizens, NYC Government Officials, and Agencies use the created and 
closed dates to measure the \textit{duration} of SRs, which often reflects an Agency's 
responsiveness. While duration is not directly present in the dataset, 
it can be easily computed as the difference between
\texttt{(closed\_date} and \texttt{created\_date}.  Duration is one of 
most frequently used 311metrics and is constantly analyzed, such as determining if 
one NYC borough receives faster response than another for certain 
complaints. There are 12,450 SRs where the \texttt{closed\_date} precedes the 
\texttt{created\_date}, generating nonsensical ``negative durations.'' 
Though this represents only 0.2\% of the dataset, these errors can 
significantly impact response time analyses. Eight SRs with extremely 
large negative durations ($-$730 days), all originating from the 
Department of Homeless Services (DHS); all contain an entry 
of ``1900-01-01'' as the \texttt{closed\_date}. This results 
in negative durations exceeding $-$44,601 days (or 122 years). Many
other SRs have negative durations in excess of 300+ days. Such 
anomalies, though rare, can significantly skew statistical results if 
not addressed during data cleansing. As a result, these SR rows are 
removed from the analysis. Here is a sample:

\begin{table}[tbp]
  \centering
  \caption{Negative duration of the largest magnitude ($-$730 days) excluding
    extreme negative values}
  \begin{tabular}{l l l r l}
    \toprule
    {created\_date} & {closed\_date} & {duration} 
    & \textbf{agency} \\
    \midrule
    2023-01-27 14:40:00 & 2022-01-14 14:40:00 & $-$378.0 & DOT \\
    2023-01-18 10:06:00 & 2022-01-12 10:06:00 & $-$371.0 & DOT \\
    2023-01-27 14:36:00 & 2022-01-22 14:35:00 & $-$370.0 & DOT \\
    2023-01-11 11:10:00 & 2022-01-09 11:10:00 & $-$367.0 & DOT \\
    2023-12-18 03:13:00 & 2023-01-16 13:10:00 & $-$335.6 & DOT \\
    \bottomrule
  \end{tabular}
  \label{tab:largest-errors}
\end{table}

@@

\subparagraph{Identical Created Date \& Closed Date: Zero Durations}
A more prevalent issue occurs when the \texttt{closed\_date} and 
\texttt{created\_date} are exactly the same, down to the second. This 
creates a \textbf{zero duration} for the SR, which is nonsensical. There 
are 163,720 such SRs, representing 2.6\% of all non-blank data. 
99\% percent of these zero-duration SRs occur within five agencies:
DHMH, DOT, DOB, DSNY, and DEP. This distribution \textbf{does not} 
mirror the overall SR distribution by Agency, suggesting 
an agency-specific issue.

@@

These unusual patterns of SR creation and closure exactly at midnight 
and noon suggest the presence of a bulk create/close software process 
that automatically assigns time stamps of midnight (00:00:00) or noon 
(12:00:00) to large batches of SRs. This behavior distorts the 
calculated duration of these SRs by providing inaccurate dates. The 
distribution by agency shows that over 92\% of the ``closed-exactly-at-midnight'' 
SRs come from just two agencies: DOB (69\%) and DSNY  (23\%).

\subparagraph{Resolution Action Update Date}
A similar issue occurs when the \texttt{resolution\_action\_update\_date} occurs well after the
\texttt{closed\_date}.  When an SR is updated, the 311 software 
automatically populates the \texttt{resolution\_action\_update\_date}. 
While it is certainly possible, and even routine, to update an SR after it is closed, 
some of these updates seem to be well beyond an expected 
date, with some ``updates'' occurring over 2 years later. As shown in 
Figure~\ref{fig:resolution-violin}, there are 10,265 SRs that 
were updated >30 days after the 
\texttt{closed\_date} (but <730 days, an arbitrary cutoff used to 
exclude infeasible dates such as ``1900-01-01''. The chart 
highlights the distribution of these post-closure updates, with 
a notable concentration of updates occurring within 
the 30-to-90-day range. This pattern raises questions about whether 
such delayed updates are standard operational practice or indicative 
of a potential issue which might warrant further investigation.

\begin{figure}[tbp]
  \centering
  \includegraphics[width=\textwidth]{post_closed_violin_chart.pdf}
  \caption{Post-closed \texttt{resolution\_action\_update\_date}(s) >30 days.}
  \label{fig:resolution-violin}
\end{figure}

@@

An question of precision vs. accuracy arises with the \texttt{latitude} 
and \texttt{longitude} fields. Both Latitude are expressed as 
a 14-decimal number, e.g. 40.86769186022511. Given 
that 1 degree of latitude at the equator is equal to 111.04 
kilometers, the ``1'' at the end of that number represents 
approximately 1.1 nanometers (1/1,000,000,000 of a meter). For 
reference a DNA molecule is approximately 2nm in width. The 
representation of the \texttt{latitude} and \texttt{longitude} fields appears to be a 
classic case of extreme precision, but limited accuracy.

@@

\texttt{Latitude/Longitude} and \texttt{location}: The \texttt{location} field 
is a pure concatenation of the \texttt{latitude} and \texttt{longitude} fields except with a 
comma and parentheses added. This makes the \texttt{location} field 
arguably harder to use than the individual fields. For example:  
\texttt{latitude}: 40.768456429488, \texttt{longitude}: $-$73.9575661888774, 
form the \texttt{location}: (40.768456429488, $-$73.95756618887745).

\texttt{borough} \& \texttt{park\_borough}: These two fields are fully redundant, 
as they are 100\% matches.

\texttt{borough} \& \texttt{taxi\_company\_borough}: Despite their names, these 
fields are almost entirely different, with only a 0.05\% match. The 
\texttt{taxi\_company\_borough} field is used exclusively by the TLC, 
indicating a need for consultation with TLC to understand the differing 
uses. It is likely that that the usage of \texttt{taxi\_company\_borough}
is quite different than that of \texttt{borough}. Such information could be 
useful only when studying TLC related SRs. It might make more
sense to store TLC data in a separate dataset.
